A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
have new data, will process now. also put some stuff in where the conclusion should be

%
%
%
%
%
%
\section{Discussion}

\AT{Short section to recap the analysis results, discuss trends, consolidate our suggestions, talk about argument vs. function poly.}

@@

\AT{DRAFT}

In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system which is reasonable given its coverage of existing code.

We will look at argument-level polymorphism, and then function level polymorphism.

@@

@@

%
%
%
%
\section{Related Work}
\AT{Not sure what the best place for this is, but we should discuss these papers somewhere.}
The idea of measuring polymorphism via some program analysis isn't new.
For instance, some work~\cite{aakerblom2015measuring} ... .
More generally, there is an existing literature on analyzing usage patterns
of language features.  For instance, the dynamic features of Smalltalk were
analyzed in some work~\cite{callau2011howdevelopers}, ...
Other analysis work~\cite{milojkovic2017duck} ... .
\section{Conclusions and Future Work}
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
\end{document}

%
% L1+ Stuff? We don't need this anymore.
% 
% 
%
%At this point, we have seen a number of reasonable polymorphic patterns that arise in R, both in L0 and L1.
%The next type system design we consider will deal with these, and we refer to it as L$1^{+}$.
%The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsL1p} and~\ref{tab:funcountsL1p} respectively.
%
%% latex table generated in R 3.5.1 by xtable 1.8-3 package
%% Fri Mar 22 18:03:30 2019
%\begin{table}[ht]
%\label{tab:argcountsL1p}
%\centering
%\begin{tabular}{lrr}
%  \hline
% Argument Polymorphism & Count & Percentage \\
%  \hline
%  Monomorphic in Type & 869K & 95.01 \\
%%  Monomorphic in Class & 831K & 90.31 \\
%%  Monomorphic in Attribute Pattern & 830K & 90.25 \\
%  Total Seen & 914K & --- \\
%     \hline
%\end{tabular}
%\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
%\end{table}
%
%% latex table generated in R 3.5.1 by xtable 1.8-3 package
%% Fri Mar 22 18:10:46 2019
%\begin{table}[ht]
%\label{tab:funcountsL1p}
%\centering
%\begin{tabular}{lrr}
%  \hline
% Function Polymorphism & Count & Percentage \\
%  \hline
%  Monomorphic in Type & 185K & 85.30 \\
%%  Monomorphic in Class & 163K & 75.13 \\
%%  Monomorphic in Attribute Pattern & 161K & 74.31 \\
%  Total Seen & 216K & --- \\
%   \hline
%\end{tabular}
%\caption{Account of {\it function} polymorphism in L$1^{+}$.}
%\end{table}
%
%We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
%To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
%See Table~\ref{tab:toppolyL1p}.
%
%% latex table generated in R 3.5.1 by xtable 1.8-3 package
%% Tue Mar 26 17:46:45 2019
%\begin{table}[ht]
%\label{tab:toppolyL1p}
%\centering
%\begin{tabular}{lrr}
%  \hline
%Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\
%  \hline
%  matrix, \realv & 6360 & 13.94 \\
%  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\
%  \C, \realv & 2774 & 6.08 \\
%  \sC, \reals & 2552 & 5.59 \\
%  data.frame, matrix & 2167 & 4.75 \\
%  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\
%  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\
%  data.frame, \realv & 1046 & 2.29 \\
%  \sL, \reals & 920 & 2.02 \\
%  \sC, \realv & 868 & 1.90 \\
%   \hline
%\end{tabular}
%\caption{Top 10 argument signatures in L$1^{+}$.}
%\end{table}
%
%A few of the top polymorphic signatures stand out.
%First, the matrix/\realv signature seems interesting, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
%In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
%\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
%v <- c(1, 2, 3, 4)
%ma <- matrix(data=v, nrow=4, ncol=1)
%v[2] == m[2] == m[2, 1]
%
%mb <- matrix(data=v, nrow=2, ncol=2)
%v[3] == m[3] == m[1, 2]
%\end{lstlisting}}\caption{Using a matrix like a vector.}\label{fig:matasvec}\end{figure}
%
%This similarity is reminiscent of another pattern which was more clear in Table~\ref{tab:L0toppoly}; that of list/vector polymorphism.
%Transforming between vectors and matrices is just as easy as transforming between lists and vectors is simple.
%\AT{We should figure out how we want to deal with this polymorphism.}

>>>>

%
%  ls trace_export_promises > promise_packages_FINAL.txt
%
%
%
%
\section{Discussion}

@@

In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system for R which is supported by empirical evidence that it captures the patterns of existing R code.
We will look at argument-level polymorphism, and then function level polymorphism.

@@

To determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
There may also be other interesting per-function trends which our current analysis does not pick up on.
We aim to make our data available for this purpose.

@@

%
%
%
%
\section{Related Work}

The idea of measuring polymorphism via some program analysis isn't new.
For instance, some work~\cite{aakerblom2015measuring} ... .

More generally, there is an existing literature on analyzing usage patterns
of language features.  
For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, ...

Other analysis work~\cite{milojkovic2017duck} ... .

%
%. nohup parallel -j55 -a promise_reduce_me.txt Rscript mar_new_processing_files/promise_processed_reduced_to_new_form.R &
%
%
%
%
\section{Conclusions and Future Work}

Ours is a community which strives to improve the lives of programmers, and understanding programmers is key to ensure that our vision aligns with their needs.
Increasingly, programming is becoming accessible to a wider audience, with many domain-specific languages being developed to address the concerns of specialists such as analysts, mathematicians, and data scientists.
Designing systems and languages to meet the specific needs of these domain experts can prove to be challenging, as our own intuitions as (often) classically-trained programmers may fail us, and in many cases they know their own needs better than we do.

One particularly interesting language which falls into this category, and indeed the object of this work, is R.
R is a language made by statisticians, for statisticians. 
The language itself is fascinating: it is a vectorized, dynamically
typed, lazy, and functional, with limited side-effects, extensive
reflective facilities and retrofitted object-oriented programming support.
Understanding how these disparate and sometimes adversarial features interact is critical to improving or otherwise further developing the language.

In our work, we built up an understanding of R by analyzing the usage of types across a large corpus of R code.
With this knowledge, we highlighted some interesting patterns of type usage, and developed a number of type systems with differing levels of granularity to describe and capture the use of types in the language.
This task was complicated by the many notions of type in R, as well as the lack of non-package R code, but we persevered:
leveraging the \genthat tool, we analyzed tests, examples, and vignettes of said packages as a close approximation for how these packages would in fact be used by programmers.
Our final type system, {\bf L2}, consolidated information from 3 of R's main reflective functions (\code{typeof}, \code{class}, and \code{attributes}), as well as some extra information about certain values (e.g. the type of list elements, and the length of vectorized types) to be as precise as reasonable.

We found that a type system such as {\bf L2} captures an overwhelming majority (>90\%) of argument-level type usage, and in it a vast majority ($\approx$80\%) of functions were monomorphic.
Among other interesting data points, we discovered that many argument positions with a vectorized type in R are monomorphic {\it and strictly scalar}, suggesting that types or annotations for scalar-only values is useful.

\AT{Need a somewhat stronger conclusion.}

%
%
%
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
\end{document}
